---
title: "Differential Gene expression analysis using Deseq2"
author: "Nitin Beesabathuni"
date: "`r Sys.Date()`"
format: 
  gfm:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
---

In this analysis, we walk through a differential gene expression workflow using DESeq2 to study transcriptional changes under hypoxic conditions and to evaluate the effects of OC2 overexpression in LNCaP cells and OC2 knockdown in PC3 cells. The goal is not only to identify differentially expressed genes, but also to demonstrate how to evaluate data quality, interpret model behavior, and assess biological plausibility at each step.

```{r}
#| results: false
#| message: false
# Load necessary library
library(tidyverse)
library(here)
library(conflicted)
library(DESeq2)
library(ggplot2)
library(ggpubr) 
library(scales) # For readable axis labels
library(gt)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(patchwork)
library(ggplotify)
library(patchwork)
library(org.Hs.eg.db)
# Resolve namespace conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflicts_prefer(generics::setdiff)

#loading functions
source("functions.R")
```

## 1. Data Import and Metadata Preparation

We begin by importing an annotated gene-level count matrix, using gene symbols as primary identifiers. Sample metadata are constructed to explicitly encode treatment condition and genetic perturbation status, as these variables will later define the design matrix used by DESeq2. Careful metadata preparation at this stage is critical, as all downstream modeling relies on the correctness of these annotations.

```{r}
# Loading cleaned and annotated data for both cell types
data <- read.csv("analysis_ready_count_data.csv", row.names = 1, check.names = FALSE)

#Separting data for both cell types into their respective dfs

data_lncap <- data %>%
  select(contains("LNCaP"))

data_pc3 <- data %>%
  select(contains("PC3"))

# 1. Sanitize the column names of your dataframes immediately
# This replaces dashes or illegal characters with dots (e.g., rep-1 becomes rep.1)
colnames(data_lncap) <- make.names(colnames(data_lncap))
colnames(data_pc3) <- make.names(colnames(data_pc3))

# 2. Define the metadata vectors
condition <- c(rep("Normoxia", 4), rep("Hypoxia", 4))
overexpress <- c(rep("Empty_vector", 2), rep("OC2_Overexpress", 2), 
                 rep("Empty_vector", 2), rep("OC2_Overexpress", 2))
knockdown <- c(rep("siCtrl", 2), rep("siOC2", 2), 
               rep("siCtrl", 2), rep("siOC2", 2))

# 3. Create the data frames with explicit row names from the sanitized columns
# We also use stringsAsFactors = TRUE or explicitly wrap in factor() for DESeq2
meta_lncap <- data.frame(
  Treatment = factor(condition, levels = c("Normoxia", "Hypoxia" )),
  Genetic_perturbation = factor(overexpress, levels = c("Empty_vector", "OC2_Overexpress")),
  row.names = colnames(data_lncap) 
)

meta_pc3 <- data.frame(
  Treatment = factor(condition, levels = c("Normoxia", "Hypoxia")),
  Genetic_perturbation = factor(knockdown, levels = c("siCtrl", "siOC2")),
  row.names = colnames(data_pc3)
)

# CRITICAL CHECK: DESeq2 will fail if these are not identical
stopifnot(all(colnames(data_lncap) == rownames(meta_lncap)))
stopifnot(all(colnames(data_pc3) == rownames(meta_pc3)))
```

## 2. Data QC for any technical artifacts

Before performing differential expression analysis, it is essential to assess whether technical artifacts or outlier samples are present. While replicate correlations at the count level provide a useful first check, high-dimensional methods such as principal component analysis (PCA) and hierarchical clustering offer deeper insight into global sample relationships.

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-dpi: 300
##defining deseq2 object
dds_pc3<- DESeqDataSetFromMatrix(countData = data_pc3,
                              colData = meta_pc3,
                              design = ~Treatment*Genetic_perturbation)   
dds_lncap<- DESeqDataSetFromMatrix(countData = data_lncap,
                              colData = meta_lncap,
                              design = ~Treatment*Genetic_perturbation) 
# For PC3
pc3_QC<-QC_pca_heatmap(dds_pc3, meta_pc3,cell_line_name = "PC3")

# For LNCaP
lncap_QC<-QC_pca_heatmap(dds_lncap, meta_lncap,cell_line_name = "LNCaP")
```

```{r}
#| fig-width: 10
#| fig-height: 3
#| fig-dpi: 300
#PCA for both the cell lines across two treatments and genetic perturbations. 
lncap_QC$pca+ pc3_QC$pca 
```

For LNCaP samples, PCA reveals four well-separated clusters corresponding to the combination of treatment condition and genetic perturbation. The first principal component (PC1), which explains approximately 56% of the total variance, primarily captures treatment effects. The second principal component (PC2), explaining approximately 37% of the variance, reflects the impact of OC2 overexpression. The absence of isolated samples or unexpected clustering patterns suggests that these data are free of major technical artifacts and suitable for downstream modeling.

PC3 samples exhibit a different structure. PCA shows a strong separation between siCtrl and siOC2 conditions; however, within the siOC2 group, samples further separate into two distinct clusters- not based on the treatment. This suggesting the presence of an additional latent variable. To investigate this further, lets examine sample correlation using heatmaps.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-dpi: 300
#heatmap for pc3 cells
pc3_QC$heatmap
```

Based on the heatmap, we see another latent variable- the siRNA used. We can see the siRNA specific samples cluster more strongly together regardless of the treatment. If treatment effects dominate global expression variance, we would expect the siRNA1 and siRNA2 to cluster together based on the treatment. However, we see a stronger correlation between the same siRNA's used for knockdown of OC2 regardless of the treatment. This could perhaps be a result of knockdown of OC2 which rescues some aspects of hypoxia driven cellular changes (not fully). However, we also see the siRNA's under normoxia cluster separately so there could be some off target activity of the specific siRNA used to knockdown OC2.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-dpi: 300
#heatmap for lncap cells
lncap_QC$heatmap
```

For LNCaP cells, the replicates cluster together based on the treatment and genetic perturbation as expected. Next, lets perform differential expression analysis using Deseq2 to ask various questions.

## 3. Differential expression analysis using Deseq2

With data quality assessed, we proceed to differential expression analysis using DESeq2. Reference levels are set such that normoxia and control genetic perturbations (empty vector or siCtrl) serve as baseline conditions. Calling the `DESeq()` function initiates a sequence of modeling steps, which are outlined below to clarify how DESeq2 arrives at its final statistical tests.

```{r}
#
#releveling to set reference for treatment and perturbation
dds_pc3$Treatment <- relevel(dds_pc3$Treatment, ref = "Normoxia")
dds_pc3$Genetic_perturbation <- relevel(dds_pc3$Genetic_perturbation, ref = "siCtrl")
  
dds_lncap$Treatment <- relevel(dds_lncap$Treatment, ref = "Normoxia")
dds_lncap$Genetic_perturbation <- relevel(dds_lncap$Genetic_perturbation, ref = "Empty_vector")

dds_lncap <- DESeq(dds_lncap)
dds_pc3<-DESeq(dds_pc3)
```

You can see various steps that are performed by this function. Briefly, the DESeq function performs the following steps-

1\) **estimating size factors**- Estimate sample-specific size factors using the median-of-ratios method to normalize for sequencing depth and composition bias.

2\) **estimating dispersions/gene-wise dispersion estimates**- Estimate gene-wise dispersion parameters under a negative binomial model.

3\) **mean-dispersion relationship**- Fit a mean–dispersion trend across genes.

4\) **final dispersion estimates**-Calculate the final dispersion estimates using shrinkage toward the fitted trend.

5\) **fitting model and testing**- Fit a negative binomial generalized linear model using the specified design matrix and perform hypothesis testing (Using Wald test here) to identify differentially expressed genes.

Now, lets examine the mean-dispersion plot to check how well our assumptions hold and

```{r}
#| fig-width: 12
#| fig-height: 5
#| fig-dpi: 300
# Set the graphics device to 1 row and 2 columns
par(mfrow = c(1, 2))

# Plot for LNCaP with a custom title
plotDispEsts(dds_lncap, main = "Dispersion Estimates: LNCaP")

# Plot for PC3 with a custom title
plotDispEsts(dds_pc3, main = "Dispersion Estimates: PC3")
```

The fit looks good. We can see the fitted (Red) smoothly follows the data points and shrinkage in dispersion (blue dots) towards fitted curve nicely. Also, notice a nice downward trend in dispersion as mean increases. We can also observe some final dispersion and estimated values perfectly overlap, these are genes that are not shrunk towards the fitted line.

```{r}
# Reset the graphics device to default (1x1)
par(mfrow = c(1, 1))

#lets understand the genes that are differential expressed under hypoxia condition. 
# PC3: Hypoxia vs Normoxia (within siCtrl)
res_pc3_hypoxia <- results(dds_pc3, 
                           contrast = c("Treatment", "Hypoxia", "Normoxia"))

# LNCaP: Hypoxia vs Normoxia (within empty vector)
res_lncap_hypoxia <- results(dds_lncap, 
                             contrast = c("Treatment", "Hypoxia", "Normoxia"))

```

Before focusing on individual differentially expressed genes (DEGs), lets also xamine the relationship between normalized counts and estimated fold changes to ensure that model estimates behave sensibly across the expression range.

```{r}
#| fig-width: 10
#| fig-height: 5
#| fig-dpi: 300
# 1. Prepare the grid (1 row, 2 columns)
par(mfrow = c(1, 2))

# 2. Call the plots directly
plotMA(res_lncap_hypoxia, ylim = c(-4, 4), alpha = 0.05,main = "LNCaP: Hypoxia vs Normoxia (Empty Vector)")
plotMA(res_pc3_hypoxia, ylim = c(-4, 4), alpha = 0.05,main = "PC3: Hypoxia vs Normoxia (siCtrls)")

# 3. Reset to default (1x1)
par(mfrow = c(1, 1))
```

Overall, we identify a larger number of DEGs in LNCaP cells compared to PC3 cells. This difference may reflect biological differences between the cell lines, differences in variability across samples, or latent technical factors.

## 4. Interpreting Differential Expression Results

```{r}
#| message: false
#| fig-width: 20
#| fig-height: 8
#| fig-dpi: 300

source("functions.R")
# Generate plots for both cell lines
v_pc3 <- plot_volcano(res_pc3_hypoxia, "PC3: Hypoxia vs Normoxia (siCtrl)")
v_lncap <- plot_volcano(res_lncap_hypoxia, "LNCaP: Hypoxia vs Normoxia (Empty vector)")

# Display side-by-side
 v_lncap+v_pc3
```

We can see some a good number of genes that are differential expressed. For PC3 samples, we can see some well known genes regulated under hypoxic conditions such as ANGPTL4 and NDRG1. Lets do a sanity check for the differential expressed genes are consistent with the known biology. Lets perform pathway enrichment on these DEGs to understand the pathways regulated. For additional confirmation at a global scale of all the DEGs, lets perform pathway enrichment analysis to identify if the known pathways are identified.

```{r}
#| message: false
#| fig-width: 15
#| fig-height: 5
#| fig-dpi: 300
# extract 
source("functions.R")
get_sig_genes <- function(res) {
  # Filter for padj < 0.05 and |LFC| > 1
  sig <- subset(res, padj < 0.05 & abs(log2FoldChange) > 1)
  
  # Convert Gene Symbols (rownames) to Entrez IDs
  ids <- bitr(rownames(sig), fromType = "SYMBOL", 
              toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  return(ids$ENTREZID)
}

genes_pc3 <- get_sig_genes(res_pc3_hypoxia)
genes_lncap <- get_sig_genes(res_lncap_hypoxia)

pc3_kegg<-plot_kegg_pathways(genes_pc3, "PC3: Hypoxia Regulated Pathways", p_cutoff = 0.05, max_pathways = 10)

# For LNCaP
lncap_kegg<-plot_kegg_pathways(genes_lncap, "LNCaP: Hypoxia Regulated Pathways", p_cutoff = 0.05, max_pathways = 10)

pc3_kegg+lncap_kegg
```

We observe Glycolysis and HIF-1 signaling pathways are some of the pathways thar regulated under hypoxia conditions. This provides additional confirmation that the analysis is correct.

```{r}

```
